---
output:
  rmarkdown::github_document:
    html_preview: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.path = "man/figures/README-", fig.align = "center",
                      fig.height = 4.5, fig.width = 4.5)
```

<!--- README.md is generated from README.Rmd. Please edit that file -->

## wdpar: Interface to the World Database on Protected Areas

[![lifecycle](https://img.shields.io/badge/Lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R-CMD-check-Ubuntu](https://img.shields.io/github/workflow/status/prioritizr/wdpar/Ubuntu/master.svg?label=Ubuntu)](https://github.com/prioritizr/wdpar/actions)
[![R-CMD-check-Windows](https://img.shields.io/github/workflow/status/prioritizr/wdpar/Windows/master.svg?label=Windows)](https://github.com/prioritizr/wdpar/actions)
[![R-CMD-check-Mac-OSX](https://img.shields.io/github/workflow/status/prioritizr/wdpar/Mac%20OSX/master.svg?label=Mac%20OSX)](https://github.com/prioritizr/wdpar/actions)
[![Coverage Status](https://codecov.io/github/prioritizr/wdpar/coverage.svg?branch=master)](https://codecov.io/github/prioritizr/wdpar?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/wdpar)](https://CRAN.R-project.org/package=wdpar)

```{r, include = FALSE}
devtools::load_all()
```

```{r, include = FALSE}
# define helpers
is_rel_version <- function(x) {
  length(strsplit(as.character(x), ".", fixed = TRUE)[[1]]) == 3
}

package_version_year <- function(x) {
  if (x %in% rownames(available.packages())) {
    return(live_package_version_year(x))
  } else {
    return(archive_package_version_year(x))
  }
}

live_package_version_year <- function(x) {
  # fetch data
  d <- versions::available.versions(x)[[1]]
  d <- d[which(d$version == max(d$version)), , drop = FALSE]
  c(as.character(d$version), as.character(format(as.POSIXct(d$date), "%Y")))
}

archive_package_version_year <- function(x) {
  # fetch data (copied from remotes R package)
  tf <- tempfile(fileext = ".gz")
  on.exit(unlink(tf), add = TRUE)
  download.file(tf, url = sprintf("%s/src/contrib/Meta/archive.rds", repo))
  con <- gzfile(tf, "rb")
  on.exit(close(con), add = TRUE)
  d <- readRDS(con)[[x]]
  # extract version
  rel_version <- basename(rownames(d))
  rel_version <- sapply(strsplit(rel_version, "_", fixed = TRUE), `[[`, 2)
  rel_version <- sapply(strsplit(rel_version, ".tar", fixed = TRUE), `[[`, 1)
  rel_version <- max(rel_version)
  # extract year
  rel_year <- as.character(max(format(d[, "mtime"], "%Y")))
  # return year version
  c(rel_version, rel_year)
}
```

### Overview

The [World Database on Protected Areas (WDPA)](https://www.protectedplanet.net/en) is the most comprehensive global dataset of protected areas. It is used to monitor the performance of existing protected areas and pinpoint priority areas for establishing new protected areas. Additionally, this database receives monthly updates from government agencies and non-governmental organizations. However, this dataset is associated with [many issues that need to be addressed prior to analysis](https://www.protectedplanet.net/en/resources/calculating-protected-area-coverage) and the dynamic nature of this dataset means that the entire data cleaning process needs to be repeated after obtaining a new version of the dataset. The _wdpar R_ package provides an interface to the [World Database on Protected Areas (WDPA)](https://www.protectedplanet.net/en) and functions for cleaning the data following best practices (outlined in [Butchart _et al._ 2015](https://doi.org/10.1111/conl.12158); [Runge _et al._ 2015](https://www.science.org/doi/abs/10.1126/science.aac9180); [Protected Planet](https://www.protectedplanet.net/en/resources/calculating-protected-area-coverage)).

### Installation

#### Package

The [latest official version of the _wdpar R_ package](https://CRAN.R-project.org/package=wdpar) can be installed using the following R code. Please note that this package requires the [_curl_](https://github.com/jeroen/curl) and [_sf R_](https://github.com/r-spatial/sf) packages which may require additional software to be installed. If you encounter problems installing the _wdpar R_ package, please consult the installation instructions for these packages.

```{r, eval = FALSE}
install.packages("wdpar", repos = "https://cran.rstudio.com/")
```

Alternatively, the latest developmental version can be installed using the following _R_ code. Please note that while developmental versions may contain additional features not present in the official version, they may also contain coding errors.

```{r, eval = FALSE}
if (!require(remotes))
  install.packages("remotes")
remotes::install_github("prioritizr/wdpar")
```

#### Additional dependencies

The _wdpar R_ package can leverage the _prepr R_ package to augment data cleaning procedures. Since the _prepr R_ package is not available on the Comprehensive R Archive Network, it is listed as an optional dependency. In some cases, the _prepr R_ package is required to complete the data cleaning procedures (e.g. to fix especially extreme geometry issues) and the _wdpar R_ package will throw an error if the package is not available. To install the _prepr R_ package, please use the following R code.

```{r, eval = FALSE}
if (!require(remotes))
  install.packages("remotes")
remotes::install_github("dickoa/prepr")
```

Note that the _prepr R_ package has system dependencies that need to be installed before the package itself can be installed (see below for platform-specific instructions).

##### _Windows_

The [Rtools](https://cran.r-project.org/bin/windows/Rtools/) software needs to be installed to install the _prepr R_ package package from source. This software provides system requirements from [rwinlib](https://github.com/rwinlib/).

##### _Ubuntu_

The `gmp`, `mpfr`, and several spatial libraries need to be installed. For recent versions of Ubuntu (18.04 and later), these libraries are available through official repositories. They can be installed using the following system commands:

```
apt-get -y update
apt-get install -y libgmp3-dev libmpfr-dev libudunits2-dev libgdal-dev libgeos-dev libproj-dev
```

##### _Linux_

For Unix-alikes, `gmp` (>= 4.2.3), `mpfr` (>= 3.0.0), and `gdal` (>= 3.2.2) are required.

##### _MacOS_

The `gmp`, `mpfr`, and `gdal` libraries are required. The easiest way to install these libraries is using [HomeBrew](https://brew.sh/). After installing HomeBrew, these libraries can be installed using the following commands in the system terminal:

```
brew install pkg-config
brew install gmp
brew install mpfr
brew install gdal
```

### Usage

Here we will provide a short introduction to the _wdpar R_ package. First, we will load the _wdpar R_ package. We will also load the _dplyr_ and _ggmap R_ packages to help explore the data.

```{r, message = FALSE, warning = FALSE}
# load packages
library(wdpar)
library(dplyr)
library(ggmap)
```

Now we will download protected area data for Malta from [Protected Planet](https://www.protectedplanet.net/en). We can achieve this by specifying Malta's country name (i.e. `"Malta"`) or Malta's [ISO3 code](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) (i.e. `"MLT"`). Since data are downloaded to a temporary directory by default, we will specify that the data should be downloaded to a persistent directory. This means that R won't have to re-download the same dataset every time we restart our R session, and R can simply re-load previously downloaded datasets as needed.

```{r, include = FALSE}
# download protected area data for Malta
mlt_raw_pa_data <- wdpa_fetch("Malta", wait = TRUE, force = TRUE)
```

```{r, eval = FALSE}
# download protected area data for Malta
mlt_raw_pa_data <- wdpa_fetch("Malta", wait = TRUE,
                              download_dir = rappdirs::user_data_dir("wdpar"))
```

Next, we will clean the data set. Briefly, the cleaning steps include: excluding protected areas that are not yet implemented, excluding protected areas with limited conservation value, replacing missing data codes (e.g. `"0"`) with missing data values (i.e. `NA`), replacing protected areas represented as points with circular protected areas that correspond to their reported extent, repairing any topological issues with the geometries, and erasing overlapping areas. For more information, see `?wdpa_clean`.

```{r}
# clean Malta data
mlt_pa_data <- wdpa_clean(mlt_raw_pa_data)
```

Print preview of the data associated with each protected area.

```{r}
# print preview
head(mlt_pa_data)
```

Finally, after cleaning the data, let's plot a map showing Malta's protected areas and color each area according to its management category ([as defined by the The International Union for Conservation of Nature](https://www.iucn.org/theme/protected-areas/about/protected-area-categories)).

```{r "readme-map", message = FALSE, warning = FALSE, out.width = ifelse(isTRUE(knitr:::is_html_output(excludes = c("markdown"))), "60%", "50%")}
# reproject data to longitude/latitude for plotting
mlt_pa_data <- st_transform(mlt_pa_data, 4326)

# download basemap imagery
bg <- get_stamenmap(unname(st_bbox(mlt_pa_data)), zoom = 8,
                    maptype = "watercolor", force = TRUE)

# make map
ggmap(bg) +
geom_sf(aes(fill = IUCN_CAT), data = mlt_pa_data, inherit.aes = FALSE) +
theme(axis.title = element_blank(), legend.position = "bottom")
```

If you need to calculate protected area coverage statistics for a country, please note that you will need to manually clip the cleaned protected area data to the countries' coastline and its Exclusive Economic Zone (EEZ) to obtain accurate results (see [official data cleaning guidelines](https://www.protectedplanet.net/en/resources/calculating-protected-area-coverage)). This step is not performed by the _wdpar R_ package because there is no single "best" coastline and Exclusive Economic Zone (EEZ) dataset, since the "best" dataset for any given project depends on the level of required precision and available computational resources. For more examples---including an example of clipping the cleaned data to a coastline---please refer to the [package vignette](https://prioritizr.github.io/wdpar/articles/wdpar.html).

### Citation

```{r, include = FALSE}
dev_version_year <- c(
  as.character(packageDescription("wdpar")$Version),
  format(Sys.time(), "%Y")
)
if (is_rel_version(dev_version_year[[1]])) {
  rel_version_year <- dev_version_year
} else {
  rel_version_year <- package_version_year("wdpar")
}
```

Please cite the _wdpar R_ package and the World Database on Protected Areas (WDPA) in publications.

To cite the latest official version, please use:

```{r, echo = FALSE, results = "asis", comment = ""}
cat(paste0("> Hanson JO (", rel_version_year[[2]],"). wdpar: Interface to the World Database on Protected Areas. R package version ", rel_version_year[[1]],". Available at https://CRAN.R-project.org/package=wdpar.\n"))
```

To cite the latest development version, please use:

```{r, echo = FALSE, results = "asis", comment = ""}
cat(paste0("> Hanson JO (", dev_version_year[[2]],"). wdpar: Interface to the World Database on Protected Areas. R package version ", dev_version_year[[1]],". Available at https://github.com/prioritizr/wdpar.\n"))
```

To cite the World Database on Protected Areas (WDPA), please use:

```{r, echo = FALSE, results = "asis", comment = ""}
cat(paste0("> UNEP-WCMC and IUCN ([insert year of the version downloaded]) Protected Planet: The World Database on Protected Areas (WDPA), [insert month/year of the version downloaded], Cambridge, UK: UNEP-WCMC and IUCN. Available at: www.protectedplanet.net."))
```
