---
title: "wdpar: Interface to the World Database on Protected Areas"
author: "Jeffrey O. Hanson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: false
    fig_caption: true
    self_contained: true
fontsize: 11pt
documentclass: article
vignette: >
  %\VignetteIndexEntry{wdpar}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r, include = FALSE}
h = 3.5
w = 3.5
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(fig.align = "center", eval = !is_check,
                      root.dir = normalizePath("../.."),
                      fig.height = 4.5, fig.width = 4.5)
```

```{r, include = FALSE}
devtools::load_all()
```

### Overview

The _wdpar R_ package provides an interface to the [World Database on Protected Areas (WDPA)](https://www.protectedplanet.net/). It provides functions for downloading and cleaning data following best practices (outlined in [Butchart _et al._ 2015](https://dx.doi.org/10.1111/conl.12158); [Runge _et al._ 2015](https://dx.doi.org/10.1126/science.aac9180); and [Protected Planet](https://www.protectedplanet.net/c/calculating-protected-area-coverage)).

### Tutorial

Here we will provide a short introduction to the _wdpar R_ package. First, we will load the _wdpar R_ package. We will also load the _sf_ and _dplyr R_ packages to help explore the data. Additionally, we will load the _ggmap R_ package to visualize the data.

```{r, message = FALSE, warning = FALSE}
# load packages
library(wdpar)
library(dplyr)
library(ggmap)
```

Now we will download protected area data for Malta from [Protected Planet](protectedplanet.net). Note that we could have alternatively downloaded the data using Malta's [ISO3 code](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) (MLT).

```{r}
# download protected area data for Malta
mlt_raw_pa_data <- wdpa_fetch("Malta")
```

Next, we will clean the data set. See the help page for `wdpa_clean` for a detailed description on the data cleaning process.

```{r}
# clean Malta data
mlt_pa_data <- wdpa_clean(mlt_raw_pa_data)
```

Now that we have finished cleaning the data, let's preview the data. For more information on what these columns mean, please refer to the [official manual](https://www.protectedplanet.net/c/wdpa-manual) (available in [English](http://wcmc.io/WDPA_Manual), [Spanish](http://wcmc.io/WDPA_Manual_ES), [French](http://wcmc.io/WDPA_Manual_FR), and [Russian](http://wcmc.io/WDPA_Manual_RU)).

```{r}
# print first six rows of the data
head(mlt_pa_data)
```

We will now reproject the data to longitude/latitude coordinates ([EPSG:4326](http://spatialreference.org/ref/epsg/wgs-84/)) for visualization purposes.

```{r}
# reproject data
mlt_pa_data <- st_transform(mlt_pa_data, 4326)
```

Next, we can plot a map showing the boundaries of Malta's protected area system.

```{r,  message = FALSE, warning = FALSE}
# download basemap for making the map
bg <- get_stamenmap(unname(st_bbox(mlt_pa_data)), zoom = 8,
                    maptype = "watercolor", force = TRUE)

# print map
ggmap(bg) +
geom_sf(data = mlt_pa_data, fill = "#31A35480", inherit.aes = FALSE) +
theme(axis.title = element_blank())
```

We can also create a histogram showing the year when each protected area was established.

```{r}
hist(mlt_pa_data$STATUS_YR, main = "Malta's protected areas", xlab = "Year established")
```

Now let's calculate some statistics. We can calculate the total amount of land and ocean inside Malta's protected area system (km^2^).

```{r}
# calculate total amount of area inside protected areas (km^2)
statistic <- mlt_pa_data %>%
             as.data.frame() %>%
             select(-geometry) %>%
             group_by(MARINE) %>%
             summarize(area_km = sum(AREA_KM2)) %>%
             ungroup() %>%
             arrange(desc(area_km))

# print statistic
print(statistic)
```

We can also calculate the percentage of land inside its protected area system that are managed under different categories (i.e. [using the protected area management categories defined by The International Union for Conservation of Nature](https://www.iucn.org/theme/protected-areas/about/protected-area-categories)).

```{r}
# calculate percentage of land inside protected areas (km^2)
statistic <- mlt_pa_data %>%
             as.data.frame() %>%
             select(-geometry) %>%
             group_by(IUCN_CAT) %>%
             summarize(area_km = sum(AREA_KM2)) %>%
             ungroup() %>%
             mutate(percentage = (area_km / sum(area_km)) * 100) %>%
             arrange(desc(area_km))

# print statistic
print(statistic)
```

We can also plot a map showing Malta's protected areas and color each area according to it's management category.

```{r, message = FALSE, warning = FALSE}
ggmap(bg) +
geom_sf(aes(fill = IUCN_CAT), data = mlt_pa_data, inherit.aes = FALSE) +
theme(axis.title = element_blank(), legend.position = "bottom")
```

### Citation

Please cite the _wdpar R_ package and the World Database on Protected Areas (WDPA) in publications.

```{r, echo  = FALSE, result = "asis", comment = ""}
citation("wdpar")
```
