---
title: "wdpar: Interface to the World Database on Protected Areas"
author: "Jeffrey O. Hanson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: false
    fig_caption: true
    self_contained: true
fontsize: 11pt
documentclass: article
vignette: >
  %\VignetteIndexEntry{wdpar: Interface to the World Database on Protected Areas}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r, include = FALSE}
h = 3.5
w = 3.5
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(fig.align = "center", eval = !is_check, purl = !is_check,
                      root.dir = normalizePath("../.."),
                      fig.height = 4.5, fig.width = 4.5)
```

```{r, include = FALSE}
devtools::load_all()
```

### Overview

The [World Database on Protected Areas (WDPA)](https://www.protectedplanet.net/en) is the most comprehensive global dataset of protected areas. It is used to monitor the performance of existing protected areas and pinpoint priority areas for establishing new protected areas. Additionally, this database receives monthly updates from government agencies and non-governmental organizations. However, this dataset is associated with [many issues that need to be addressed prior to analysis](https://www.protectedplanet.net/en/resources/calculating-protected-area-coverage) and the dynamic nature of this dataset means that the entire data cleaning process needs to be repeated after obtaining a new version of the dataset. The _wdpar R_ package provides an interface to the [World Database on Protected Areas (WDPA)](https://www.protectedplanet.net/en) and functions for cleaning the data following best practices (outlined in [Butchart _et al._ 2015](https://dx.doi.org/10.1111/conl.12158); [Runge _et al._ 2015](https://www.sciencemag.org/doi/abs/10.1126/science.aac9180); [Protected Planet](https://www.protectedplanet.net/en/resources/calculating-protected-area-coverage)).

### Tutorial

Here we will provide a short introduction to the _wdpar R_ package. First, we will load the _wdpar R_ package. We will also load the _dplyr_ and _ggmap R_ packages to help explore the data.

```{r, message = FALSE, warning = FALSE}
# load packages
library(wdpar)
library(dplyr)
library(ggmap)
```

Now we will download protected area data for Malta from [Protected Planet](https://www.protectedplanet.net/en). We can achieve this by specifying Malta's country name (i.e. `"Malta"`) or Malta's [ISO3 code](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) (i.e. `"MLT"`). Since data are downloaded to a temporary directory by default, we will specify that the data should be downloaded to a persistent directory. This means that R won't have to re-download the same dataset every time we restart our R session, and R can simply re-load previously downloaded datasets as needed.

```{r}
# download protected area data for Malta
# (excluding areas represented as point localities)
mlt_raw_pa_data <- wdpa_fetch("Malta", wait = TRUE,
                              download_dir = rappdirs::user_data_dir("wdpar"))
```

Next, we will clean the data set. Briefly, the cleaning steps include: excluding protected areas that are not yet implemented, excluding protected areas with limited conservation value, replacing missing data codes (e.g. `"0"`) with missing data values (i.e. `NA`), replacing protected areas represented as points with circular protected areas that correspond to their reported extent, repairing any topological issues with the geometries, and erasing overlapping areas. For more information, see `?wdpa_clean`.

```{r}
# clean Malta data
mlt_pa_data <- wdpa_clean(mlt_raw_pa_data)
```

After cleaning the data set, we will perform an additional step that involves
clipping the terrestrial protected areas to Malta's coastline. Ideally, we would
also clip the marine protected areas to Malta's Exclusive Economic Zone (EEZ) but such data are not as easy to obtain on a per country basis (but see https://www.marineregions.org/eez.php)).

```{r}
# download Malta boundary from Global Administrative Areas dataset
file_path <- tempfile(fileext = "rds")
download.file("https://biogeo.ucdavis.edu/data/gadm3.6/Rsf/gadm36_MLT_0_sf.rds",
              file_path)

# import Malta's boundary
mlt_boundary_data <- readRDS(file_path)

# repair any geometry issues, dissolve the border, reproject to same
# coordinate system as the protected area data, and repair the geometry again
mlt_boundary_data <-
  mlt_boundary_data %>%
  st_set_precision(1000) %>%
  sf::st_make_valid() %>%
  st_set_precision(1000) %>%
  st_combine() %>%
  st_union() %>%
  st_set_precision(1000) %>%
  sf::st_make_valid() %>%
  st_transform(st_crs(mlt_pa_data)) %>%
  sf::st_make_valid()

# clip Malta's protected areas to the coastline
mlt_pa_data <-
  mlt_pa_data %>%
  filter(MARINE == "terrestrial") %>%
  st_intersection(mlt_boundary_data) %>%
  rbind(mlt_pa_data %>%
        filter(MARINE == "marine") %>%
        st_difference(mlt_boundary_data)) %>%
  rbind(mlt_pa_data %>% filter(!MARINE %in% c("terrestrial", "marine")))

# recalculate the area of each protected area
mlt_pa_data <-
  mlt_pa_data %>%
  mutate(AREA_KM2 = as.numeric(st_area(.)) * 1e-6)
```

Now that we have finished cleaning the data, let's preview the data. For more information on what these columns mean, please refer to the [official manual](https://www.protectedplanet.net/en/resources/wdpa-manual) (available in English, French, Spanish, and Russian).

```{r}
# print first six rows of the data
head(mlt_pa_data)
```

We will now reproject the data to longitude/latitude coordinates ([EPSG:4326](https://spatialreference.org/ref/epsg/wgs-84/)) for visualization purposes.

```{r}
# reproject data
mlt_pa_data <- st_transform(mlt_pa_data, 4326)
```

Next, we can plot a map showing the boundaries of Malta's protected area system.

```{r,  message = FALSE, warning = FALSE}
# download basemap for making the map
bg <- get_stamenmap(unname(st_bbox(mlt_pa_data)), zoom = 8,
                    maptype = "watercolor", force = TRUE)

# print map
ggmap(bg) +
geom_sf(data = mlt_pa_data, fill = "#31A35480", inherit.aes = FALSE) +
theme(axis.title = element_blank())
```

We can also create a histogram showing the year when each protected area was established.

```{r}
hist(mlt_pa_data$STATUS_YR, main = "Malta's protected areas", xlab = "Year established")
```

Now let's calculate some statistics. We can calculate the total amount of land and ocean inside Malta's protected area system (km^2^).

```{r}
# calculate total amount of area inside protected areas (km^2)
statistic <-
  mlt_pa_data %>%
  as.data.frame() %>%
  select(-geometry) %>%
  group_by(MARINE) %>%
  summarize(area_km = sum(AREA_KM2)) %>%
  ungroup() %>%
  arrange(desc(area_km))

# print statistic
print(statistic)
```

We can also calculate the percentage of land inside its protected area system that are managed under different categories (i.e. [using the protected area management categories defined by The International Union for Conservation of Nature](https://www.iucn.org/theme/protected-areas/about/protected-area-categories)).

```{r}
# calculate percentage of land inside protected areas (km^2)
statistic <-
  mlt_pa_data %>%
  as.data.frame() %>%
  select(-geometry) %>%
  group_by(IUCN_CAT) %>%
  summarize(area_km = sum(AREA_KM2)) %>%
  ungroup() %>%
  mutate(percentage = (area_km / sum(area_km)) * 100) %>%
  arrange(desc(area_km))

# print statistic
print(statistic)
```

We can also plot a map showing Malta's protected areas and color each area according to it's management category.

```{r, message = FALSE, warning = FALSE}
ggmap(bg) +
geom_sf(aes(fill = IUCN_CAT), data = mlt_pa_data, inherit.aes = FALSE) +
theme(axis.title = element_blank(), legend.position = "bottom")
```

### Recommended practices for large datasets

The _wdpar R_ package can be used to clean large datasets -- such as those spanning large countries, multiple countries, and even the full global datatset -- assuming that sufficient computational resources and time are available. For example, when processing the global dataset, it is recommended to use a computer system with at least 32 GB RAM available and to allow for at least one full day for the data cleaning procedures to complete. It is also recommended to avoid using the computer system for any other tasks while the data cleaning procedures are being completed, because they are very computationally intensive. Additionally, when processing large datasets -- and especially for the global dataset -- it is strongly recommended to disable the procedure for erasing overlapping areas. This is because the built-in procedure for erasing overlaps is very time consuming when processing many protected areas, so that information on each protected area can be output (e.g. IUCN category, year established). Instead, it is recommended to run the data cleaning procedures with the procedure for erasing overlapping areas disabled (i.e. with `erase_overlaps = FALSE`). After the data cleaning procedures have completed, the resulting protected area data can be subject to additional criteria for refining the dataset if needed (e.g. if you only want protected areas that are highly suitable for conserving biodiversity, you may wish to exclude protected areas that are not categorized by the IUCN as Ia or Ib). Finally, to obtain a set of protected area boundaries with no overlapping areas, the protected area data can be dissolved together (using `wdpa_dissolve()`). For an example of these procedures, please see below.

```{r}
# download protected area data for multiple of countries
## (i.e. Portugal, Spain, France)
raw_pa_data <-
  c("PRT", "ESP", "FRA") %>%
  lapply(wdpa_fetch, wait = TRUE,
         download_dir = rappdirs::user_data_dir("wdpar")) %>%
  bind_rows()

# clean protected area data (with procedure for erasing overlaps disabled)
full_pa_data <- wdpa_clean(raw_pa_data, erase_overlaps = FALSE)

# subset protected area data based on additional criteria
## (e.g. retain only protected areas that are IUCN category Ia and Ib)
sub_pa_data <-
  full_pa_data %>%
  filter(IUCN_CAT %in% c("Ia", "Ib"))

# dissolve all geometries together (removing spatial overlaps)
pa_data <- wdpa_dissolve(sub_pa_data)

# preview data
print(pa_data)

# calculate total area
print(st_area(pa_data))
```

### Additional datasets

Although the [World Database on Protected Areas (WDPA)](https://www.protectedplanet.net/en) is the most comprehensive global dataset, many datasets are available for specific countries or regions that do not require such extensive data cleaning procedures. As a consequence, it is often worth looking for alternative data sets when working at smaller geographic scales before considering the [World Database on Protected Areas (WDPA)](https://www.protectedplanet.net/en). The list below outlines several alternative protected area datasets and information on where they can be obtained. If you know of any such datasets that are missing, [please create an issue on the GitHub repository](https://github.com/prioritizr/wdpar/issues) and we can add them to the list.

* Arctic
    + [Arctic Protected Areas - 2017](http://geo.abds.is/geonetwork/srv/eng/catalog.search#/metadata/2e56ee1f-50a9-4983-88f4-edaa8588950d)
* Australia
    + [Collaborative Australian Protected Area Database (CAPAD)](https://www.environment.gov.au/land/nrs/science/capad)
    + [Conservation Areas Reporting and Tracking System (CARTS)](https://ccea-ccae.org)
* The United States of America
    + [Protected Areas Database of the US, PAD-US (CBI Edition)](https://consbio.org/products/projects/pad-us-cbi-edition)

### Citation

Please cite the _wdpar R_ package and the World Database on Protected Areas (WDPA) in publications. To see citation details, use the code:

```{r, eval = FALSE}
citation("wdpar")
```
