---
title: "wdpar: Interface to the World Database on Protected Areas"
author: "Jeffrey O. Hanson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: false
    fig_caption: true
    self_contained: true
fontsize: 11pt
documentclass: article
vignette: >
  %\VignetteIndexEntry{wdpar}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r, include = FALSE}
h = 3.5
w = 3.5
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(fig.align = "center", eval = !is_check,
                      root.dir = normalizePath("../.."))
```

```{r, include = FALSE}
devtools::load_all()
```

Here we will provide a short introduction to the _wdpar R_ package.

First, we will load the _wdpar R_ package. We will also load the _sf_ and _dplyr R_ packages to help explore protected area data.

```{r, message = FALSE, warning = FALSE}
# load packages
library(wdpar)
library(sf)
library(dplyr)
```

Now we will download protected area data for Malta. Note that we could have alternatively downloaded the data using Malta's [ISO3 code](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) (MLT).

```{r}
# download protected area data for Malta
mlt_raw_pa_data <- wdpa_fetch("Malta")
```

Next, we will clean the data set. Note that we will only use a single thread for data processing, but you could use the `threads` argument to use multiple threads for data processing. See the help page for `wdpa_clean` for a detailed description on the data cleaning process.

```{r}
# clean Malta data
mlt_pa_data <- wdpa_clean(mlt_raw_pa_data)
```

Now that we have finished cleaning the data, let's calculate some statistics. We can calculate the total amount of area inside inside Malta's protected area system (km$^2$). Note that this includes marine and terrestrial protected areas.

```{r}
# calculate total amount of area inside protected areas (km^2)
statistic <- mlt_pa_data %>%
             as.data.frame %>%
             summarize(area = sum(AREA_KM2))

# print statistic
print(statistic)
```

We can also calculate the percentage of land inside [protected areas that are designated strictly to protect biodiversity (and also potentially geological/geomorphological features)](https://www.iucn.org/theme/protected-areas/about/protected-area-categories).

```{r}
# fetch spatial data for Malta's land mass and calculate its area (km^2)
mlt_land_data <- land_and_eez_fetch("Malta") %>%
                 filter(TYPE == "LAND") %>%
                 summarize(area = as.numeric(st_area(.)) * 1e+6)

# calculate percentage of land inside protected areas (km^2)
statistic <- mlt_pa_data %>%
             as.data.frame %>%
             filter(MARINE == "terrestrial", IUCN_CAT == "Ia") %>%
             summarize(area_protected = sum(AREA_KM2)) %>%
             mutate(total_land_area = mlt_land_data[["area"]]) %>%
             mutate(percentage_protected = (area_protected / total_land_area) *
                                           100)

# print statistic
print(statistic)
```

Finally, let's plot a map showing Malta's protected areas and color each area according to its management category.

```{r, fig.height = 2.5, fig.width = 4.5}
plot(mlt_pa_data[, "IUCN_CAT"], main = "IUCN Category", key.size = lcm("4"))
```
